---
title: "SCRAPI2: Propagating GE uncertainty in smolt escapement estimation"
subtitle: "Validation against SCOBI::SCRAPI — MY2025 steelhead"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
    highlight: tango
vignette: >
  %\VignetteIndexEntry{SCRAPI2: GE uncertainty in smolt escapement estimation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  eval    = TRUE
)
```

---

# Background

## What SCRAPI does

SCRAPI (Steinhorst & Ackerman) estimates wild smolt escapement past Lower Granite
Dam (LGR) by expanding trap catches to the full passage using a daily sampling
rate. The core formula for daily estimated passage is:

$$\hat{N}_d = \frac{C_d}{R_d \times GE_d}$$

where $C_d$ is the trap count, $R_d$ is the sampling rate (fraction of fish
entering the bypass that are trapped), and $GE_d$ is the guidance efficiency —
the fraction of all downstream migrants that actually enter the juvenile bypass
system. The bootstrap resamples both the passage counts and the trap fish to
propagate uncertainty in $R_d$ and the stock/rear composition into 90% CIs.

## The GE problem

Historically, $GE_d$ is a **fixed point estimate** taken from a NOAA table
(typically modelled from spill and fish behaviour studies). Plugging in a fixed
$GE$ means the bootstrap CIs only reflect uncertainty in trap sampling — not
uncertainty in GE itself. For years with sparse PIT tag data or unusual spill
conditions this can substantially understate the true uncertainty in escapement.

## What SCRAPI2 adds

`SCRAPI2` extends the existing SCRAPI logic in two ways, either of which can be
used independently or together:

1. **GE draws (`geDraws`)** — instead of a single $GE_d$ per day, supply a
   matrix of posterior draws from a Bayesian hierarchical model. Each bootstrap
   iteration uses a different column (draw) so GE uncertainty flows through
   into the CIs.

2. **GSI draws (`gsiDraws`)** — instead of fixed stock assignments in the trap,
   supply posterior draws from a genetic stock identification model. Each
   bootstrap iteration uses a different draw so genetic uncertainty flows through
   into per-stock CIs.

**Key guarantee:** when every column of `geDraws` is identical (or when
`geDraws = NULL` and a fixed `GuidanceEfficiency` column is supplied), `SCRAPI2`
produces results numerically identical to `SCOBI::SCRAPI`.

---

# Install and load

```{r install, eval=FALSE}
remotes::install_github(
  "dsmi313/escapeLGD",
  ref = "claude/multi-github-pages-connection-sC5FR"
)
```

```{r libraries}
library(smoltEASE)
library(tidyverse)
library(lubridate)
library(readxl)
library(coda)      # gelman.diag for MCMC convergence
```

---

# Data inputs

## Trap data, FPC, spill, strata

```{r data-load}
# PIT tag detection history (full season)
MY25.pit <- read.csv("MY2025_full.csv")

# Daily spill (kcfs)
spill <- read.csv("spill.csv") %>%
  mutate(Date = as.Date(Date, "%m/%d/%Y"))

# Pull trap data from LGD database
sthd_raw  <- query_lgd_trap("MY2025", "3", "JV")
STHD.trap <- lgr2SCRAPI(sthd_raw, species = "sthd")

# FPC passage data (sampling rates, counts, dates)
sthd_fpc_raw <- read_fpc("2025SMPCollectionsAtLGR_CH1&ST.xlsx", species = "sthd")
sthd_fpc     <- build_fpc(sthd_fpc_raw, STHD.trap,
                           export_csv = "MY2025STHDsmolts_FPCdata.csv")

# QC
check_fpc(sthd_fpc,  "STHD")
check_trap(STHD.trap, "STHD", check_age = TRUE)
```

## Strata

Week-to-stratum collapse mapping. Each row maps one ISO week number to a
stratum integer. Weeks in the same stratum are pooled for estimation.

```{r strata}
strata_raw  <- read_excel("Smolt Strata.xlsx", sheet = "Sheet1",
                           col_names = FALSE)
sthd_start  <- which(strata_raw[[1]] == "WeekNumber")[2]

sthd_strata <- strata_raw[(sthd_start + 1):nrow(strata_raw), c(1, 3)] %>%
  setNames(c("Week", "Collapse")) %>%
  filter(!is.na(Week)) %>%
  mutate(Week = as.integer(Week), Collapse = as.integer(Collapse))

# Optional: visualise stock composition by week to inform stratum choice
genstock_by_week(STHD.trap, species = "sthd")
```

## Fixed GE table (NOAA/legacy)

```{r fixed-ge}
ge_sthd <- read_excel("GE 2025.xlsx",
                       sheet = "Wild Steelhead 260406", skip = 23) %>%
  select(DOY = 1, GuidanceEfficiency = 3) %>%
  filter(!is.na(DOY), !is.na(GuidanceEfficiency)) %>%
  mutate(SampleEndDate = as.Date("2024-12-31") + as.integer(DOY)) %>%
  select(SampleEndDate, GuidanceEfficiency)
```

---

# The Bayesian GE model

## What it estimates

Fish passing LGR take one of two measurable routes: the juvenile bypass (detected
at GRS) or through spillway/turbines without bypass contact (detected at GRJ or
not detected at LGR at all). From PIT-tagged fish we can estimate:

$$\psi_s = P(\text{detected at GRS} \mid \text{passed LGR via bypass})$$

per stratum $s$. This is the **spillway detection probability** — fish that enter
the bypass but miss the detector are counted as undetected (UND). Then:

$$GE_s = \frac{n_{\text{GRJ},s}}{n_{\text{GRJ},s} + n_{\text{GRS},s} / \hat\psi_s}$$

The model links $\psi_s$ to spill via a hierarchical logistic regression:

$$\text{logit}(\psi_s) \sim \mathcal{N}(\alpha + \beta \cdot \text{spill}_s,\ \sigma_\psi^2)$$

with weakly informative priors $\alpha, \beta \sim \mathcal{N}(0, 10)$ and
$\sigma_\psi \sim \text{HalfCauchy}(0, 1)$. Fitting is done in JAGS via
`fit_ge_model()`. Each MCMC draw of $\{\psi_s\}$ becomes one column in `geDraws`,
so GE uncertainty propagates through every bootstrap iteration of SCRAPI2.

## Prepare the stratum-level GE data

```{r ge-prep}
dat_up  <- prep_pit_data(MY25.pit, species = "Steelhead", mig_year = 2025)

ge_data <- prep_ge_data(
  dat_up,
  strat_assign = sthd_strata,
  spill_data   = spill,
  species      = "sthd"
)

# Stratum-level summary used by the model
ge_data %>%
  mutate(GE_empirical = round(n_GRJ_obs / (n_GRJ_obs + n_GRS_obs / psi), 3)) %>%
  select(stratum, spill_kcfs = spill_val,
         n_GRJ = n_GRJ_obs, n_GRS = n_GRS_obs, psi, GE_empirical) %>%
  knitr::kable(caption = "Stratum-level inputs to the GE model")
```

## Fit the model and check convergence

```{r ge-fit}
set.seed(123)
ge_fit <- fit_ge_model(ge_data)

# Gelman-Rubin R-hat: values < 1.1 indicate convergence
gelman.diag(ge_fit$samples)
```

## Compare model GE to NOAA fixed GE

```{r ge-compare-plot}
# Aggregate NOAA fixed GE to stratum level (weekly mean within each collapse)
noaa_by_collapse <- ge_sthd %>%
  mutate(Week = isoweek(SampleEndDate)) %>%
  semi_join(sthd_strata, by = "Week") %>%
  left_join(sthd_strata, by = "Week") %>%
  group_by(Collapse) %>%
  summarise(NOAA_GE = mean(GuidanceEfficiency, na.rm = TRUE), .groups = "drop")

plot_dat <- ge_data %>%
  mutate(raw_GE = n_GRJ_obs / (n_GRJ_obs + n_total_GRS)) %>%
  left_join(noaa_by_collapse, by = c("stratum" = "Collapse")) %>%
  select(stratum, spill_kcfs = spill_val, Model_psi = psi, Raw_GE = raw_GE, NOAA_GE) %>%
  pivot_longer(c(Model_psi, Raw_GE, NOAA_GE),
               names_to = "source", values_to = "GE")

ggplot(plot_dat, aes(x = spill_kcfs, y = GE, colour = source, label = stratum)) +
  geom_line(aes(group = source), linewidth = 0.8) +
  geom_point(size = 3) +
  geom_text(nudge_y = 0.015, show.legend = FALSE, size = 3) +
  scale_colour_manual(
    values = c(Model_psi = "#2166ac", Raw_GE = "#d73027", NOAA_GE = "#4dac26"),
    labels = c("Bayesian model (psi)", "Raw empirical GE", "NOAA fixed GE")
  ) +
  labs(title  = "GE estimates by stratum and method",
       x      = "Mean spill (kcfs)",
       y      = "Guidance efficiency",
       colour = NULL) +
  theme_minimal(base_size = 13)
```

## Generate GE draws for SCRAPI2

```{r ge-draws}
B <- 5000

ge_draws_sthd <- generate_ge_draws(
  ge_fit       = ge_fit,
  strat_assign = sthd_strata,
  pass_dates   = sthd_fpc$SampleEndDate,
  B            = B
)

# Sanity checks
stopifnot(nrow(ge_draws_sthd) == nrow(sthd_fpc))
stopifnot(all(unlist(ge_draws_sthd[, -1]) >= 0))
stopifnot(all(unlist(ge_draws_sthd[, -1]) <= 1))

cat("Posterior mean GE by day:\n")
summary(rowMeans(ge_draws_sthd[, -1]))
```

---

# Three-way comparison

We run three estimation workflows on the same MY2025 steelhead data and compare
the resulting wild smolt estimates and confidence intervals.

| Workflow | GE source | Tool |
|---|---|---|
| **A — Legacy** | NOAA fixed table | `SCOBI::SCRAPI` |
| **B — SCRAPI2 fixed GE** | Same NOAA table, no uncertainty | `SCRAPI2` |
| **C — SCRAPI2 model GE** | Bayesian posterior draws | `SCRAPI2` |

Paths A and B should produce nearly identical point estimates — this validates
that `SCRAPI2` has not changed the underlying SCRAPI estimation logic. Path C
should have wider CIs because GE uncertainty is now properly propagated.

## Helper: add Collapse column for legacy SCOBI::SCRAPI

`SCOBI::SCRAPI` requires the `Collapse` column to be pre-baked into both data
frames. `SCRAPI2` handles this internally via its `strata =` argument.

```{r add-collapse-helper}
add_collapse <- function(df, date_col, strata_tbl) {
  df %>%
    mutate(.d   = as.Date(!!sym(date_col)),
           Week = isoweek(.d)) %>%
    left_join(strata_tbl, by = "Week") %>%
    mutate("{date_col}" := format(.d, "%Y-%m-%d")) %>%
    select(-.d)
}
```

## Path A — Legacy `SCOBI::SCRAPI`

```{r path-a-legacy}
STHD.trap_legacy <- add_collapse(STHD.trap, "CollectionDate", sthd_strata)

sthd_fpc_legacy  <- sthd_fpc %>%
  left_join(ge_sthd, by = "SampleEndDate") %>%
  add_collapse("SampleEndDate", sthd_strata)

stopifnot(!anyNA(STHD.trap_legacy$Collapse))
stopifnot(!anyNA(sthd_fpc_legacy$Collapse))

SCOBI::SCRAPI(
  smoltData   = STHD.trap_legacy,
  passageData = sthd_fpc_legacy,
  Primary     = "GenStock",
  dateFormat  = "%Y-%m-%d",
  Run         = "legacy_",
  B           = B
)

# SCOBI::SCRAPI returns NULL and writes output to CSV; read it back
legacy_ci <- read.csv("legacy_CIs.csv", skip = 1,
                       header = TRUE, row.names = 1, check.names = FALSE)
```

## Path B — `SCRAPI2` with fixed GE (validation run)

```{r path-b-fixed-ge}
sthd_fpc_fixed_ge <- left_join(sthd_fpc, ge_sthd, by = "SampleEndDate")

result_fixed <- SCRAPI2(
  smoltData   = STHD.trap,
  passageData = sthd_fpc_fixed_ge,
  strata      = sthd_strata,
  Primary     = "GenStock",
  dateFormat  = "%Y-%m-%d",
  B           = B
)
```

## Path C — `SCRAPI2` with Bayesian GE draws

```{r path-c-model-ge}
result_model <- SCRAPI2(
  smoltData   = STHD.trap,
  passageData = sthd_fpc,
  strata      = sthd_strata,
  geDraws     = ge_draws_sthd,
  Primary     = "GenStock",
  dateFormat  = "%Y-%m-%d",
  B           = B
)
```

---

# Results

## Wild smolt totals

```{r results-table}
# Unified extraction helper
ci_df <- function(mat, label) {
  as.data.frame(mat) %>%
    rownames_to_column("Stock") %>%
    mutate(Workflow = label)
}

# Paths B and C return result$CI (matrix with Estimate / LCI / UCI / P1)
# Path A is read from the CSV SCOBI wrote
tbl_legacy <- legacy_ci %>%
  rownames_to_column("Stock") %>%
  mutate(Workflow = "A: SCOBI legacy") %>%
  rename(Estimate = Estimate, LCI = LCI, UCI = UCI)

tbl_fixed  <- ci_df(result_fixed$CI, "B: SCRAPI2 fixed GE")
tbl_model  <- ci_df(result_model$CI, "C: SCRAPI2 model GE")

results_all <- bind_rows(tbl_legacy, tbl_fixed, tbl_model)

# Wild smolt total row only
results_all %>%
  filter(Stock == "WildSmolts") %>%
  mutate(`CI half-width (%)` = round((UCI - LCI) / (2 * Estimate) * 100, 1)) %>%
  select(Workflow, Estimate, LCI, UCI, `CI half-width (%)`) %>%
  knitr::kable(
    caption = "Wild steelhead smolt estimates — three workflows",
    format.args = list(big.mark = ",")
  )
```

## Forest plot — all stocks

```{r forest-plot, fig.width=9, fig.height=5}
results_all %>%
  filter(Stock != "WildSmolts") %>%
  mutate(Stock = factor(Stock)) %>%
  ggplot(aes(x = Estimate, xmin = LCI, xmax = UCI,
             y = Stock, colour = Workflow)) +
  geom_pointrange(position = position_dodge(width = 0.5), size = 0.6) +
  scale_colour_manual(values = c(
    "A: SCOBI legacy"      = "#555555",
    "B: SCRAPI2 fixed GE"  = "#2166ac",
    "C: SCRAPI2 model GE"  = "#d73027"
  )) +
  labs(title  = "Wild steelhead smolts by genetic stock — MY2025",
       x      = "Estimated abundance",
       y      = NULL,
       colour = NULL) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")
```

## CI width comparison — WildSmolts

```{r ci-width-plot, fig.width=6, fig.height=4}
results_all %>%
  filter(Stock == "WildSmolts") %>%
  mutate(
    half_width = (UCI - LCI) / 2,
    Workflow   = fct_inorder(Workflow)
  ) %>%
  ggplot(aes(x = Workflow, y = half_width, fill = Workflow)) +
  geom_col(width = 0.5, show.legend = FALSE) +
  scale_fill_manual(values = c(
    "A: SCOBI legacy"      = "#555555",
    "B: SCRAPI2 fixed GE"  = "#2166ac",
    "C: SCRAPI2 model GE"  = "#d73027"
  )) +
  labs(title = "90% CI half-width — WildSmolts",
       x = NULL, y = "Half-width (fish)") +
  theme_minimal(base_size = 13)
```

The model-GE CI (Path C) should be wider than the fixed-GE CI (Paths A/B).
The difference represents the additional uncertainty from not knowing GE exactly.

---

# Control check (exact-GE draws)

This section verifies the `geDraws` pathway itself is neutral when every draw
contains the same GE value as the fixed table. Point estimates should agree to
within bootstrap noise; CI width should be identical.

```{r control-exact-ge, results='hold'}
ge0 <- pmin(pmax(as.numeric(sthd_fpc_fixed_ge$GuidanceEfficiency), 1e-6), 1 - 1e-6)

ge_draws_exact <- data.frame(
  SampleEndDate = sthd_fpc_fixed_ge$SampleEndDate,
  # Repeat the same GE value B times per row — column names don't matter,
  # SCRAPI2 uses positional access (geDraws[, -1])
  matrix(rep(ge0, B), nrow = length(ge0), ncol = B),
  check.names = FALSE
)

result_exact <- SCRAPI2(
  smoltData   = STHD.trap,
  passageData = sthd_fpc_fixed_ge,
  strata      = sthd_strata,
  geDraws     = ge_draws_exact,
  Primary     = "GenStock",
  dateFormat  = "%Y-%m-%d",
  B           = B
)

# Point estimates must agree to within bootstrap noise (< 1%)
est_fixed <- result_fixed$CI["WildSmolts", "Estimate"]
est_exact <- result_exact$CI["WildSmolts", "Estimate"]
cat(sprintf("Fixed GE estimate : %s\n", format(est_fixed, big.mark = ",")))
cat(sprintf("Exact-draw estimate: %s\n", format(est_exact, big.mark = ",")))
cat(sprintf("Difference        : %.2f%%\n",
            abs(est_fixed - est_exact) / est_fixed * 100))

stopifnot(abs(est_fixed - est_exact) / est_fixed < 0.01)
```

---

# Reviewer checklist

| Check | Expected result |
|---|---|
| `gelman.diag()` R-hat values | < 1.1 for all parameters (`psi[s]`, `alpha`, `beta`, `sigma_psi`) |
| GE draws range | All values in \[0, 1\] |
| Path A vs Path B point estimates | Within ~5% (bootstrap noise with same seed) |
| Path C CI half-width vs Path B | Wider or equal — never narrower |
| Exact-GE control check | < 1% difference in WildSmolts estimate |

---

# Session info

```{r session-info, eval=TRUE}
sessionInfo()
```
